---
title: "Exploratory Analysis on Prosper Loan Data"
author: "Sharad Mohan Vijalapuram"
date: "June 10, 2017"
output:
  html_document: default
  pdf_document: default
---
```{r set-options, include=FALSE, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(out.width='900px', dpi=200, echo = FALSE, 
                      message=FALSE, warning=FALSE)
```


## Introduction
Peer-to-peer(P2P) lending (and borrowing) is a new method of financing that has become immensely popular with in the small amount of time it has been around for. Prosper is the first P2P company in the US and now manages more than $7 billion dollars in loan amounts and is one of the leading P2P lending companies. 

I've used a subset of Prosper's loan data to analyze different variables, and provide my observations and reflection.

## Univariate Analysis

### Structure of the dataset
There are 113,937 loans in the dataset with 81 features (LoanOriginalAmount, LoanOriginationDates, BorrowerRate, LenderYield, etc., being some of them).

```{r load packages}
#Gets and loads all the packages needed for this analysis

packagesneeded <- c("ggplot2", "dplyr", "RColorBrewer", "plotly", "GGally")

for (package in packagesneeded) 
    {
        if(!(package %in% installed.packages()))
        {
        install.packages(package)
        }
    library(package, character.only = T)
    
    }  

```

```{r load data} 

pl <- read.csv('prosperLoanData.csv')

dim(pl)

str(pl)

```

By looking at the structure of the data, I observed that the total number of records in the data set and the number of listing keys don't match which could mean there are duplicates in the data set.

```{r duplicates, echo=TRUE} 
dup_pl<- pl[duplicated(pl$ListingKey) | duplicated(pl$ListingKey, fromLast=TRUE),]

length(dup_pl)

```

The length of the duplicates' data frame seems to make sense as it adds up to the total number of records in the original data set. I still want to take a closer look at the duplicates.

What I found interesting is that the data of these duplicate records are the same except for the ProsperScore field, which are different.

```{r , echo=TRUE} 
group_by(dup_pl, ListingKey) %>%
  summarise(.,duplicates = n()) %>%
  ggplot(aes(duplicates)) +
    geom_bar() +
    #scale_y_sqrt() +
    geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
    theme(text = element_text(size=8.5))

```

It looks like the ListingKey was not only repeated twice but upto 6 times for a particular loan. 3 times for about 32 of them and 4 times for 4 loans.

Let's look at the prosper scores or analyze how they could be assigned a little later.

```{r}

#Converting the feature, Term to a factor
pl$Term <- as.factor(pl$Term)

#Converting LoanOriginationDate as Date
pl$LoanOriginationDate <- as.Date(pl$LoanOriginationDate, tz = "UTC")

pl$ListingCreationYear <- as.factor(format(as.Date(pl$ListingCreationDate), '%Y'))

pl$CumulativeLoanAmount <- NA

pl[order(pl$LoanOriginationDate),]$CumulativeLoanAmount <- cumsum(pl[order(pl$LoanOriginationDate),"LoanOriginalAmount"])

pl$CumulativePayments <- NA

pl[order(pl$LoanOriginationDate),]$CumulativePayments <- 
  cumsum(pl[order(pl$LoanOriginationDate),"LP_CustomerPayments"])

pl$ListingCategory..numeric. <- factor(pl$ListingCategory..numeric.)

ListingCategory <- c('Not Available', 'Debt Consolidation', 'Home Improvement',
                     'Business', 'Personal Loan', 'Student Use', 'Auto', 'Other',
                     'Baby & Adoption', 'Boat', 'Cosmetic Procedure', 
                     'Engagement Ring', 'Green Loans',  'Household Expenses',
                     'Large Purchases',  'Medical/Dental', 'Motorcycle', 'RV',
                     'Taxes', 'Vacation', 'Wedding Loans')

pl$ListingCategory <- ListingCategory[pl$ListingCategory..numeric.]

#For use in Bivariate analysis
pl$MainLoanStatus <- ifelse(pl$LoanStatus == "Completed" | 
                            pl$LoanStatus == "Current" |
                            pl$LoanStatus == "FinalPaymentInProgress" , 
                            "Paid/Current",
                            "Unpaid/Delayed")

pl$MeanCreditScore <- (pl$CreditScoreRangeUpper + pl$CreditScoreRangeLower)/2

#Necessary variables for correlation plot
variables <- c("BorrowerAPR", "LenderYield", 
               "DebtToIncomeRatio", "StatedMonthlyIncome", 
               "LoanOriginalAmount", "MeanCreditScore", "EstimatedLoss", 
               "EstimatedReturn", "CurrentDelinquencies", "AmountDelinquent",
               "BankcardUtilization", "Investors")

#Calculating actual returns
pl$ActualReturn <- ((pl$LP_CustomerPayments - pl$LoanOriginalAmount) 
                    / pl$LoanOriginalAmount)

#Good/Bad Returns
pl$Returns <- if_else(pl$ActualReturn>0, "Good","Bad")

#State Names
statenames <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", 
                "Colorado", "Connecticut", "Delaware", "Florida", "Georgia", 
                "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", 
                "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts",
                "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana",
                "Nebraska", "Nevada", "New Hampshire", "New Jersey", 
                "New Mexico", "New York", "North Carolina", "North Dakota", 
                "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", 
                "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah",
                "Vermont", "Virginia", "Washington","West Virginia", "Wisconsin",
                "Wyoming")

#Re-ordering the factors on loan quarters
pl$LoanOriginationQuarter <- factor(pl$LoanOriginationQuarter, 
                                    levels = c("Q4 2005","Q1 2006", "Q2 2006", 
                                               "Q3 2006", "Q4 2006", "Q1 2007", 
                                               "Q2 2007", "Q3 2007", "Q4 2007", 
                                               "Q1 2008", "Q2 2008", "Q3 2008", 
                                               "Q4 2008", "Q1 2009", "Q2 2009", 
                                               "Q3 2009", "Q4 2009", "Q1 2010", 
                                               "Q2 2010", "Q3 2010", "Q4 2010", 
                                               "Q1 2011", "Q2 2011", "Q3 2011", 
                                               "Q4 2011", "Q1 2012", "Q2 2012", 
                                               "Q3 2012", "Q4 2012", "Q1 2013", 
                                               "Q2 2013", "Q3 2013", "Q4 2013", 
                                               "Q1 2014", "Q2 2014", "Q3 2014", 
                                               "Q4 2014"))

#Re-ordering the factors on income range
pl$IncomeRange <- factor(pl$IncomeRange, 
                         levels = c("Not displayed", "Not employed", "$0", 
                                    "$1-24,999", "$25,000-49,999", 
                                    "$50,000-74,999", "$75,000-99,999", 
                                    "$100,000+"))

```
### Univariate Plots

```{r, echo = TRUE}
#Creating a function to plot some basic invariate plots
basicuniplots <- function(df, v, xlab, ylab = "Number of Listings") {
  
  ggplot(aes(v), data = df) + 
    geom_bar(stat = 'Count') +
    labs (x = xlab, y = ylab) +
    theme(text = element_text(size=8.5))
  
}  
```

Since I'll end up plotting a lot of bar plots in this section, I created a function to plot them.

```{r univariate plots}
summary(pl$LoanOriginalAmount)

#Plotting Loan Original Amount
ggplot(aes(LoanOriginalAmount), data = pl) +
  geom_histogram(stat = 'bin') +
  labs (x = "Loan Original Amount", y = "Number of Listings") +
  theme(text = element_text(size=8.5))


ggplot(aes(LoanOriginalAmount), data = pl) +
  geom_histogram(stat = 'bin', binwidth = 200) +
  scale_x_continuous(limits = c(0, 36000),
                     breaks = seq(0, 36000, 5000)) +
  scale_y_sqrt(limits = c(0, 15000),
                     breaks = seq(0, 15000, 5000)) +
  theme(text = element_text(size=8.5))
#+
#  geom_vline(aes(xintercept=mean(LoanOriginalAmount, na.rm=T)),   # Ignore NA values for mean
#              color="blue", linetype="dashed", size=1)
```

The minimum loan amount taken out is $1000 and the maximum is $35,000. Although the loan range is wide, most of the loans are below $10,000 with the mean and median sitting at $8337 and $6500 respectively. 

```{r}

ggplot(aes(LoanOriginationDate), data = pl) + 
  geom_histogram(stat = 'bin', binwidth = 50) +
  labs (x = "Loan Origination Date", y = "Number of Listings") +
  theme(text = element_text(size=8.5))
```

The data set contains loan details right from the year 2006 and the company seems to have had a decent growth till late 2008 when the loans given out stalled for close to an year and resumed mid to late 2009. I wonder if the loan original amount will have any insights for the gaps in time.

```{r}
ggplot(aes(LoanOriginationDate, CumulativeLoanAmount/1000000), data = pl) + 
  geom_line() + 
  scale_y_continuous(breaks = c(0,250,500,750,950)) +
  theme(legend.position="none") +
  labs (x = "Loan Origination Date",y = "Total Loan Amount in Millions") +
  theme(text = element_text(size=8.5))
```

I have created a new variable here called CumulativeLoanAmount that cumulatively adds the LoanOriginalAmount based on the LoanOriginationDate. The total loan amount given out through Prosper is about $950 Million. It is interesting to see that the company has had a nice growth right from the beginning with a linear growth till late 2008 and an exponential growth since 2010. It is also interesting to see a flat patch during 2009 and is in sync with the previous plot.

Some online research around this period threw some light on the fact that Prosper Marketplace had changed its business model during that period from a variable rate model, with rates set by lenders and borrowers to a pre-set rate model with Propser using its own formula in determining a borrower's rate based on their creditworthiness.

```{r}
ggplot(aes(LoanOriginationDate, CumulativePayments/1000000), data = pl) + 
  geom_line() + 
  #scale_color_manual(values="Blue")
  theme(legend.position="none") +
  labs (x = "Loan Origination Date",y = "Total Payments in Millions") +
  theme(text = element_text(size=8.5))
```

The very next detail I wanted to look at is how much Prosper is getting back on thse loans. The plot above is a cumulative customer payments plot and seems to be in sync with the total loan amounts. 

```{r}
summary(pl$ProsperRating..Alpha.)

basicuniplots(pl, pl$ProsperRating..Alpha., "Prosper Rating")
```
There are quite a few loans with no prosper rating and I think these could be from the time when Prosper was on their old business model.

```{r, echo=TRUE}
ggplot(data = pl[pl$ProsperRating..Alpha. == "",], aes(LoanOriginationDate)) + 
  geom_histogram(stat = "bin", binwidth = 20) +
  labs (x = "Loan Origination Date",y = "Number of Listings") +
  theme(text = element_text(size=8.5))

```
Above is a plot to confirm the same. I wonder if they still had the credit scores on file for these borrowers and it looks like they have most of their credit scores recorded and just did not calculate their Prosper Rating.
```{r, echo=TRUE}
unique(pl[pl$ProsperRating..Alpha. == "","CreditScoreRangeLower"])
```

```{r, message=TRUE}

summary(pl$Term)

basicuniplots(pl, pl$Term, "Loan Term")
```
While Prosper offers loans at 3 different loan terms, the most opted for loan term is the 36-month loan term. Although this feature is a discrete feature in the current business model, the data set stores it as a continuous feature. To make the plot look better and for future use in other plots, this feature is converted to a factor with 3 levels. 

```{r, message=TRUE}
summary(pl$LoanStatus)

basicuniplots(pl, pl$LoanStatus, "Loan Status") + 
  coord_flip()
```
It looks like most of the loans are in good standing as a huge portion of the loans are either current or completed. I have created another feature that simply notes if a loan is in good standing or not based on its loan status (assuming current and completed loans to be good).

### Let us look at some borrower details

```{r, message=TRUE}
basicuniplots(pl, pl$BorrowerState, "Borrower State") +
  coord_flip() 
```
It is very clear that California, Texas and New York have the most number of listings created for loans. 

```{r, message=TRUE}
basicuniplots(pl, pl$EmploymentStatus, "Employment Status") + 
  coord_flip()
```
This plot is very useful for both borrowers and investors alike. It is important to note that borrowers with any kind of job have a higher probability of getting a loan with Prosper.

```{r, message=TRUE}
summary (pl$IsBorrowerHomeowner)

basicuniplots(pl, pl$IsBorrowerHomeowner, "Home Owner") 
```
This doesn't really seem to be a major factor in getting a loan as both home owners and renters have almost equal number of loans. Although, it would be interesting to see if this variable plays any role in deciding the interest rate/APR.
```{r, message=TRUE}

table (pl$ListingCategory)

basicuniplots(pl, pl$ListingCategory, "Listing Category") +
  coord_flip()
```
I added a new variable for the ListingCategory descriptions as the data set only has their numeric counterparts. I found the mapping available in the Variable Description document for Prosper.

Debt consolidation seems to be the #1 reason for loans on prosper.

```{r, message=TRUE}
summary(pl$BorrowerAPR*100)

ggplot(aes(BorrowerAPR*100), data = pl) + 
  geom_histogram(stat = 'bin', binwidth = 1.5) +
  labs (x = "Borrower APR",y = "Number of Listings") +
  theme(text = element_text(size=8.5))
```
The data set contains both, borrower APR and interest rate and since APR includes additional fees on top of the preset interest rate, I wanted to look at the Borrower APR instead of interest rate as this is what the borrower would end up paying. 

The plot seems to sit between an uniform curve and a bimodal curve with most of the loans having an APR between 
12% and 25%. It would be intersting to investigate more on the two big peaks at 30% and 36%. The peak could be due to the less creditworthy people at those rates.

```{r}

basicuniplots(pl, pl$IncomeRange, "Income Range") +
  coord_flip()
```
Borrowers with a decent income have the most number of loans. No surprises here.

#### What's in it for the investor?

```{r}

summary(pl$LenderYield*100)

ggplot(aes(LenderYield*100), data = pl) + 
  geom_histogram(stat = 'bin', binwidth = 1.5) +
  labs (x = "Lender Yield",y = "Number of Listings") +
  theme(text = element_text(size=8.5))
```
This looks like a great news for investors with mean lender yield at 17% and the max in this data set at 49.25%. This doesn't necessarily come without any risks. 

```{r}
summary (pl$Investors)

ggplot(aes(Investors), data = pl) + 
  geom_histogram(stat = 'bin', binwidth = 20) +
  scale_y_log10() +
  labs (x = "Investors",y = "Number of Lisitngs") +
  theme(text = element_text(size=8.5))
```
On average, a funded loan seems to have about 80-81 investors depending on the loan amount.

```{r}
summary (pl$Recommendations)

ggplot(aes(Recommendations), data = pl) + 
  geom_histogram(stat = 'bin', binwidth = 2) +
  scale_y_log10() +
  labs (x = "Recommendations",y = "Number of Lisitngs") +
  theme(text = element_text(size=8.5))
```
Very few borrowers seem to have recommendations. 

I wonder if borrowers had friends who lended them money.
```{r}
summary (pl$InvestmentFromFriendsCount)

ggplot(aes(InvestmentFromFriendsCount), data = pl) + 
  geom_histogram(stat = 'bin', binwidth = 1) +
  scale_y_log10() +
  labs (x = "Investment From Friends Count",y = "Number of Lisitngs") +
  theme(text = element_text(size=8.5))
```
Similar results here. Very few borrowers seems have benefitted from a loan by their friends.

#### Main observations from Univariate Analysis:

* There are only about 113066 listings in the data set.
* For some, Prosper has multiple prosper scores for the same listings.
* Number of loans has been constantly growing, linearly till mid 2009 followed by a sshort stall and exponentially   since then.
* Loan amounts vary from $1000 all the way up to #35,000.
* There are 3 loan terms avaialble, 12, 36 and 60 month terms with 36-month being the most popular.
* Till mid-2009 Prosper wasn't calculating borrower's credit score or rating, although it noted their credit        scores for most of them.
* Most of the loans are in good standing.
* Most number of loans were taken out for debt consolidation.
* California, Texas and New York have the most number of listings
* Most of the borrowers have a job with a decent income

## Bivariate Analysis

```{r}
#Re-ordering the Propser credit grades
pl$ProsperRating..Alpha.<-factor(pl$ProsperRating..Alpha., 
                                 levels=c("AA", "A", "B", "C", "D", "E", "HR")) 

ggplot(aes(ProsperRating..Alpha.,MeanCreditScore), data = pl) + 
  geom_boxplot() +
  labs (x = "Prosper Rating",y = "Mean Credit Score") +
  theme(text = element_text(size=8.5))
```
For ease of use, I created a variable called MeanCreditScore which gets the mean between the upper and lower ranges of the borrower's credit score. 

As it was observed earlier that for the years prior to 2010, Prosper was not calculating borrower's prosper rating, we have quite a few missing ratings.
Lets consider the current business model and filter out the records with missing prosper rating.

```{r}

ggplot(aes(ProsperRating..Alpha.,MeanCreditScore), 
       data = subset(pl,ProsperRating..Alpha.!="")) + 
  geom_boxplot() +
  labs (x = "Prosper Rating",y = "Mean Credit Score") +
  theme(text = element_text(size=8.5))
```
This makes perfect sense. Higher the credit score, better the rating except for HR rating. 

Lets add Prosper Score to this plot and analyze a little later.

```{r}
ggplot(aes(BorrowerAPR*100,MeanCreditScore), 
       data = subset(pl,MeanCreditScore>100)) + 
  geom_jitter(alpha = 0.1) +
  stat_smooth(method = 'lm', formula = y ~ poly(x,3), 
              se = FALSE) +
  labs (x = "Borrower APR(%) Yield",y = "Mean Credit Score")
```

On the whole, it could be said that the borrower APR increases as the borrower's credit score decreases, but there are a lot of outliers in the plot. While the credit score seems to be an important aspect in deciding a borrower's APR, it is definitely not the only factor involved as it could also be gathered from the rest of this analysis.

```{r}
by(pl$LenderYield*100, pl$ProsperRating..Alpha., summary)

ggplot(aes(ProsperRating..Alpha.,LenderYield*100), 
       data = subset(pl,ProsperRating..Alpha.!="")) + 
  geom_boxplot() +
  labs (x = "Prosper Rating",y = "Lender Yield (%)") +
  theme(text = element_text(size=8.5))
```
The relationship between lender yield and prosper rating looks as expected. Better the prosper rating, lower the yield and vice versa. High lender yield might also mean higher risk. More about this later.

```{r}

ggplot(aes(DebtToIncomeRatio,LenderYield), data = pl) + 
  geom_point(alpha = 0.2) +
  scale_x_sqrt() +
  labs (x = "Debt to Income Ratio",y = "Lender Yield") +
  theme(text = element_text(size=8.5))
```
There seems to be no direct relation between debt to income ratio and borrower rate.

Lets take a closer look at the plot. 

```{r}

ggplot(aes(DebtToIncomeRatio,LenderYield), data = pl) + 
  geom_jitter(alpha = 0.1) +
  scale_x_continuous(limits= c(0,1)) +
  stat_smooth(method = 'lm', formula = y ~ poly(x,2), se = FALSE) +
  labs (x = "Debt to Income Ratio",y = "Lender Yield") +
  theme(text = element_text(size=8.5))
```
To look at the bulk of the points, I limited debt to income ration to less than or equal to 1. There are numerous points spread out and there doesn't seem to be a correlation between them. A linear model with formula y~x seems to linearly increase with debt albeit with a minimal slope and a lot of points that are not anywhere close to the line.

```{r}
by(pl$DebtToIncomeRatio, pl$IncomeRange, summary)

ggplot(aes(DebtToIncomeRatio, fill = IncomeRange), data = pl) + 
  geom_density(alpha = 0.75) + 
  facet_wrap(~IncomeRange, scales = "free", ncol = 2) +
  scale_fill_brewer(palette="Greens") +
  labs (x = "Debt to Income Ratio",y = "Income Range") +
  theme(legend.position="none", text = element_text(size=8.5))
```

The income ranges with the least debt to income ratio is $100,000+ and $75,000-99,999. Although $100,000+ has a lower mean and median debt to income ratio, there seems to be very few outliers at 10.01.


```{r}
by(pl$CurrentDelinquencies, pl$ProsperRating..Alpha., summary)

ggplot(aes(CurrentDelinquencies, fill = ProsperRating..Alpha.), 
       data = subset(pl,ProsperRating..Alpha. !="")) + 
  geom_density(alpha = 0.75) + 
  scale_x_sqrt() +
  facet_wrap(~ProsperRating..Alpha., scales = "free", ncol = 2) +
  scale_fill_brewer(palette="RdYlGn", direction = -1) +
  labs (x = "Current Delinquencies", y = "Density") +
  theme(legend.position="none", text = element_text(size=8.5)) +
  guides(fill=guide_legend(title="Prosper Rating"))
``` 
As expected, there seems to be a pattern here. Borrowers with more delinquencies have poorer credit rating assigned by Prosper. And as the saying goes, "Correlation doesn't mean causation". 

```{r}

ggplot(aes(LoanOriginationQuarter, fill = MainLoanStatus), data = pl) + 
  geom_bar(stat = 'count') +
  scale_fill_brewer(palette="Set2") +
  labs (x = "Loan Origination Quarter") +
  theme(axis.text.x = element_text(angle = 90), text = element_text(size=8.5)) 
```

I've created a variable called main loan status that classifies whether a loan is performing well or not based on the variable Loan Status in the dataset. It looks like more and more loans recently are current or completed or not late in payments yet. 

```{r}
by(pl$ProsperScore, pl$MainLoanStatus, summary)

ggplot(aes( MainLoanStatus,ProsperScore, fill = MainLoanStatus ), data = pl) + 
  geom_boxplot() +
  scale_fill_brewer(palette="Set2") +
  labs (x = "Main Loan Status",y = "Prosper Score") +
  theme(text = element_text(size=8.5))
```

Prosper Score seems to be a decent measure and predictor of a loan turning good or bad. Borrowers with better prosper scores have loans in good status and vice versa.

#### Main observations from Bivariate Analysis:

* Borrower's credit scores plays a major role in deciding prosper's ratings/score.
* Borrower APR, Lender yield are closely dependant on prosper's credit rating.
* Borrower APR spikes at around 36% because of the borrowers with poor prosper credit ratings.
* Compared to the initial stages of Prosper, bad loans have gone down in the recent times.
* Debt to income ratio is the highest for unemployed borrowers and least for borrowers with income above $75,000.
* Prosper score is a decent measure of good and bad loans.

## Multivariate Analysis

```{r}
ggcorr(pl[, variables], hjust = 0.55, size = 2.5, label = T, label_round = 2, label_alpha = T)
```

The correlation matrix revealed a few surprising details. There doesn't seem to be a strong relationship between Borrower APR or Lender Yield and other variables including mean credit score. Interest rate of a borrower seems to be decided by Prosper based on a lot of borrower variables.

```{r}
ggplot(aes(EstimatedReturn,LoanOriginalAmount, color = ProsperRating..Alpha.),
       data = subset(pl, ProsperRating..Alpha.!="")) + 
  geom_point(alpha = 0.8) +
  labs (x = "Estimated Return",y = "Loan Original Amount") +
  scale_colour_brewer(palette="RdYlGn", direction = -1) +
  theme(text = element_text(size=8.5)) +
  guides(fill=guide_legend(title="Prosper Rating"))
```

The estimated return seems to be more dependant on the prosper rating than the original loan amount. Although, it is also clear that only borrowers with good credit rating get higher loan amounts.

```{r}
completedloans <- filter(pl, LoanStatus == "Completed")

completedloans[order(completedloans$LoanOriginationDate),]$CumulativeLoanAmount <- cumsum(completedloans[order(completedloans$LoanOriginationDate),"LoanOriginalAmount"])

completedloans[order(completedloans$LoanOriginationDate),]$CumulativePayments <- cumsum(completedloans[order(completedloans$LoanOriginationDate),"LP_CustomerPayments"])

ggplot(aes(x=LoanOriginationDate), data = completedloans) + 
  geom_line(aes(y = CumulativeLoanAmount/1000000, color = "Loan Amount")) + 
  geom_line(aes(y = CumulativePayments/1000000, 
                color = if_else(CumulativePayments>=CumulativeLoanAmount, 
                                "Good Customer Payments", 
                                "Poor Customer Payments"))) + 
  scale_color_manual(values=c("green", "blue", "red")) +
  #facet_wrap(~IncomeRange) +
  labs (x = "Loan Origination Date",
        y = "Total Loan amount and Payments in Millions") +
  theme( text = element_text(size=8.5)) +
  theme(legend.title=element_blank())

```

I used a subset of dataset that has 'completed' loans for this plot and as I hoped to see, the payments curve is well above the loan amount curve. But, this might not mean that all the investors have made a profit on these loans.
I wanted to look at those loans and to my surprise, I found quite a few investors that have not received any payments from customers on completed loans. Or, there could be bad or missing data in the data set.

```{r}
filter(pl,(pl$LoanStatus == "Completed" & 
           pl$CumulativePayments < pl$CumulativeLoanAmount &
           pl$Returns == "Bad")) %>%
  arrange(LP_CustomerPayments) %>% 
  select(LoanOriginationDate, LoanOriginalAmount, LP_CustomerPayments) %>%
  head
```

Here are a few listings that made 0 payments towards the loan and are still considered 'Completed'. One thing they have in common is that they all seemt to be in 2009. Either the borrowers had no intention of paying the money back or this is missing data as this is from Prosper's old business model. 

```{r}
ggplot(aes(EstimatedReturn, LoanOriginalAmount, fill = Returns), data = pl) + 
  stat_summary(fun.y = "mean", geom = "bar") +
  facet_wrap(~ProsperRating..Alpha., scales ="free") +
  labs (x = "Estimated Return",y = "Loan Original Amount") +
  theme(text = element_text(size=8.5))
```

I wanted to see if the original loan amount value made a difference to the estimated return.
I created a variable called Returns to visually represent the estimated return below 0. Estimated return has a very poor correlation with the original loan amount. 

```{r}
ggplot(aes(ActualReturn, LoanOriginalAmount, fill = Returns), 
       data = subset(pl,LoanStatus == "Completed")) + 
  stat_summary(fun.y = "mean", geom = "bar") +
  facet_wrap(~ProsperRating..Alpha., scales ="free") +
  labs (x = "Actual Return",y = "Loan Original Amount") +
  theme(text = element_text(size=8.5))
```

By looking at the two plots above, it is clear that the estimated return is not completely in sync with the actual returns lenders earn.

So, I wanted to plot both estimated and actual returns against each other.

```{r}
cor.test( x=pl$EstimatedReturn, y=pl$ActualReturn, method = "pearson")

ggplot(aes(ActualReturn, EstimatedReturn, color = Returns), 
       data = subset(pl,LoanStatus == "Completed")) + 
  geom_point(alpha = 0.75) +
  labs (x = "Actual Return",y = "Estimated Return") +
  theme(text = element_text(size=8.5))
```

The way actual returns are calculated, it only makes sense to look at the loans that have been completed.
So, in this plot, I'm only looking at the loans that have been completed and I still see some bad returns. 
I'm guessing these are from the old business model (pre-2010).


```{r}
ggplot(aes(ActualReturn, EstimatedReturn, color = Returns), data = pl) + 
  geom_point(alpha = 0.75) +
  facet_wrap(~MainLoanStatus) +
  labs (x = "Actual Return",y = "Estimated Return") +
  theme(text = element_text(size=8.5)) 
```

In the Paid/Current facet of the plot, Prosper was never right in estimating the bad loans. I'm looking at the negative x and y axis for this observation. 

Looking at the plot above, it is clear that the actual return is not even close to the estimated return. Prosper clearly is a little more conservative with their estimated return. This is definitely a good news for investors.

```{r}
cor.test(pl$CurrentDelinquencies, pl$BorrowerAPR, method=c("pearson"))

ggplot(aes(CurrentDelinquencies,BorrowerAPR*100, color = ProsperRating..Alpha.),
       data = subset(pl, ProsperRating..Alpha. !="")) + 
  geom_point(alpha = 0.8) +
  scale_x_sqrt() +
  scale_colour_brewer(palette="RdYlGn", direction = -1) +
  labs (x = "Current Delinquencies",y = "Borrower APR(%)") +
  theme(text = element_text(size=8.5)) +
  guides(fill=guide_legend(title="Prosper Rating"))
```

Since current delinquencies has a good linear relation with Prosper Rating and Prosper Rating has a good relation with borrower APR, I was expecting to see a stronger linear correlation between current delinquencies and borrower AP but, that doesn't seem to be the case with a mere 0.15 correlation.

```{r}
cor.test(pl$AmountDelinquent, pl$BorrowerAPR, method=c("pearson"))

ggplot(aes(AmountDelinquent,BorrowerAPR*100, color = ProsperRating..Alpha.),
       data = subset(pl, ProsperRating..Alpha. !="")) + 
  geom_point(alpha = 0.8) +
  scale_x_sqrt() +
  scale_colour_brewer(palette="RdYlGn", direction = -1) +
  labs (x = "Amount Delinquent",y = "Borrower APR(%)") +
  theme(text = element_text(size=8.5)) +
  guides(fill=guide_legend(title="Prosper Rating"))
```

Borrower's amount delinquent seems to have a very poor correlation with APR. Current delinquency matters but not the amount apprently in deciding a borrower's APR.

#### Main observations from Multivariate Analysis:

* Original loan amount almost has not effect on how the loan would turn out.
* For some, Prosper has multiple prosper scores for the same listings.
* There seem to be a few listings with 0 customer payments. 
* The returns at least on the completed loans have been growing along with the loan amount.
* Estimated return is not a very accurate estimate. It is a more generic estimate.
* Debt to Income ratio is higher for unemployed and lower income borrowers.
* A combination of borrower's prosper rating, mean score, income range and debt to income ratio have a strong effect on   the returns, rate of returns and risk of loss.

### Final Plots and Summary

#### Plot 1

```{r}
sumStates <- group_by(pl[nchar(as.character(pl$BorrowerState)) > 0,], 
                      BorrowerState) %>%
  summarise(., total_issued = round(sum(LoanOriginalAmount/1e6),1), 
            count = n(), avg_debt = round(mean(DebtToIncomeRatio, na.rm = T),1),
            avg_state_income = round(mean(StatedMonthlyIncome),1),
            avg_borrowerapr = round(mean(BorrowerAPR*100),1))

#sumStates$map_amnt = with(sumStates, pmin(total_issued, 500))

sumStates$hover <- with(sumStates, 
                        paste(state.name[match(BorrowerState, state.abb)], '<br>', 
                              "Total amount issued:",
                                prettyNum(total_issued, big.mark = ","),
                                'Million USD <br>',
                              "Number of Loans:", 
                                prettyNum(count, big.mark = ","), '<br>',
                              "Average Debt to Income Ratio:", 
                                prettyNum(avg_debt, big.mark = ","), '<br>',
                              "Average State Income:", 
                                prettyNum(avg_state_income, big.mark = ","), '<br>',
                              "Average Borrower APR(%):", 
                                prettyNum(avg_borrowerapr, big.mark = ",")))

g = list(scope = 'usa', projection = list(type = 'albers usa'), 
         showlakes = T, lakecolor = toRGB('white'))

plot_geo(sumStates, locationmode = 'USA-states') %>%
  add_trace(
    z = ~total_issued, text = ~hover, 
    locations = ~BorrowerState,
    color = ~total_issued, colors = 'Oranges'
  ) %>%
  colorbar(title = "in Million USD", lenmode = "fraction", len = 0.8) %>%
  layout(
    title = 'Total Loans and Average Debt across States',
    geo = g
  )
```

#### Description 1

The plot shows the total amount of loans issued distributed geographically throughout US. Since this is an interactive map, I have also added the number of listings, the average debt to income ratio, monthly income and borrower APR for each state to the hover window. 

#### Plot 2

```{r}
ggplot(aes(BorrowerAPR*100,MeanCreditScore, color = Term), 
       data = subset(pl,MeanCreditScore>100)) + 
  geom_point(alpha = 0.5) +
  stat_smooth(method = 'lm', formula = y ~ poly(x,2), 
              aes(colour = Term), se = FALSE) +
  facet_grid(~Term, scales = "free") +
  ggtitle("Impact of Credit Score on Borrower APR") +
  labs (x = "Borrower APR(%)",y = "Mean Credit Score") +
  theme(legend.position="none", text = element_text(size=8.5))
```

#### Description 2

This plot shows the relationship between the borrower's mean credit score and APR for different loan terms. I have also added a linear model for each term with formula, y = x^2. It shows that borrowers with higher credit scores have lower APR and there by interest rate and borrowers with lower credit scores have higher APR.

#### Plot 3

```{r}
ggplot(aes(DebtToIncomeRatio,LenderYield*100, color = ProsperRating..Alpha.), 
       data = subset(pl,ProsperRating..Alpha.!= "")) + 
  geom_point(alpha = 0.8) +
  scale_x_continuous(limits = c(0,1)) +
  scale_y_log10(breaks = c(10,20,30)) +
  scale_colour_brewer(palette="RdYlGn", direction = -1) +
  ggtitle("Impact of Prosper Rating and Debt to Income Ratio on Lender Yield(log10)") +
  labs (x = "Debt to Income Ratio",y = "Lender Yield(%)") +
  guides(fill=guide_legend(title="Prosper Rating")) +
  theme(legend.position="bottom", text = element_text(size=8.5))
```

#### Description 3

This plot shows the relationship between the borrower's debt to income ratio, prosper rating and lender's yield on those listings. Lender's yield is clearly dependant on both, debt to income ratio and prosper rating. The higher the debt to income ratio, worse the prosper rating and higher the lender yield and vice versa. 

## Reflection

The Prosper loan data set has 113,937 loan listings and  81 variables from Nov 2005 - March 2014. I was able to analyze around 20 variables over these 9 areas and also created a few additional variables to help in the analysis. The data is rich with information on borrower, credit, debt, investor, loan and all of their geographic data. 

One main problem that I faced with a lot of variables is overplotting. In a lot of plots, the amount of data and variation in the data overshadowed a trend that may or may not exist. Some of the aternatives to overcome this situation is to sample the data or cluster the data and find patterns or trends within the clusters. Machine learning might be able to solve or some of these problems. 
The amount and types of variables in the data set have introduced me to various styles and best practices of plotting certain variables.

There is still a lot of room for analysis and hopefully build models to predict things like:

1. Accurate estimated return based on the type of borrower than a generic estimated return
2. Accurate estimated loss
3. Which borrowers are more likely to default
4. Will the borrowers be able to pay back with a more conservative payment plan

Although, there were too many variables and more than 100k records available, considering this to be 9 years worth of data for US wide loans, it is a very small data set and having additional data could help strength some patterns, trends or relations that are not strongly visible now. 

## References

1. [Prosper Marketplace](https://www.prosper.com/)
2. [Prosper Wikipedia](https://en.wikipedia.org/wiki/Prosper_Marketplace)
3. [R Markdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
4. [Choropleth Maps in R - Plotly](https://plot.ly/r/choropleth-maps/)
5. [ggplot2 Reference](http://ggplot2.tidyverse.org/reference/)
6. [List of U.S. states](https://simple.wikipedia.org/wiki/List_of_U.S._states)

